---
title: Sex differences in hunting data from Koster et al. (2020)
author: Edward H. Hagen
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

source("analysis.R")
```

<!-- badges: start -->
<!-- badges: end -->

The code in this repo generates a plot of the hunting data in:

Koster, J., McElreath, R., Hill, K., Yu, D., Shepard Jr, G., Van Vliet, N., ... & Ross, C. (2020). The life history of human foraging: Cross-cultural and individual variation. Science Advances, 6(26), eaax9070. <https://doi.org/10.1126/sciadv.aax9070>

The data for the Koster et al. (2020) study was obtained here:

<https://github.com/rmcelreath/cchunts>

## Installation

Download the project repo, open it in RStudio, and install these `r` packages:

```
tidyverse
ggtext
cchunts # From github (see above)
```

## Generate this page

```
devtools::build_readme()
```

## Introduction

Koster et al. (2020) investigated hunting skill as a function of age. It was not a study of sex differences in hunting, but data collection was not restricted to male hunting. Some of the societies are hunter-gatherers, but many are not.

Here is the Koster et al. description of their dataset (from their SI, with my emphasis in **bold**):

>The total sample contains 1,821 individual hunter, 23,747 hunter-level outcomes, and 21,160 trips across 40 study sites. To compile the dataset, the first author searched for rel evant studies on subsistence hunting in the anthropological and biological literature, subsequently contacting authors to invite them to contribute data. The contributors submitted data in a standardized format that included variables for the biomass acquired on terrestrial hunting trips, the ages of the hunters at the time of the hunt, the duration of the trip, the hunting weaponry carried by the hunters, and the presence of dogs or assistants (the distinction between hunters and assistants was left to the discretion of contributors, who were counseled to conceptualize ”hunters” as those individuals who made active contributions to detecting and pursuing prey). Our data are restricted to hunting, and exclude gathering, because of the paucity of data on gathered plant foods. 
>
>There is tremendous imbalance in sample size across units. One site contributes only 6 trips from 2 individuals. Another contributes more than 14,000 trips from 147 individuals. Some individuals contribute only a single outcome, while others contribute dozens. **The majority of the sample comprises male hunters, with too little data on female hunters to infer generalizable sex differences.** (This does not imply that men’s production and skill is more relevant to human evolution, nor that women’s foraging skill would necessarily exhibit either the same or a different functional relationship with age.) Most sites contribute primarily cross-sectional data, while a few others exhibit impressive time series. The statistical framework is designed to make use of all these data.

## Methods

The data comprise one row per hunting trip. Each row records the quantity of meat returned in kilograms (harvest), the sex and age of each member of the hunting party, and whether the harvest reflects meat acquired only by the index hunter or the pooled quantity returned by the hunting party.

Since the purpose of my analyses was to assess sex differences, I removed the few hunting trips with mixed sex parties. I also added one additional variable: whether the data contributed from each society had at least one female coauthor on the associated paper.

## Results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Results
pct_solo <- 100*round(sum(d$group_size == 1)/nrow(d), 3)
males <- sum(hunters$Sex == "M")
pct_males <- 100*round(males/nrow(hunters), 3)
females <- sum(hunters$Sex == "F")
pct_females <- 100*round(females/nrow(hunters), 3)

mixed <- sum(d$mixed)
pct_mixed <- 100*round(mixed/nrow(d), 3)

harvest_males <- round(sum(d_onesex$harvest[d_onesex$sex == "M"]) / 1000, 1)
harvest_females <- round(sum(d_onesex$harvest[d_onesex$sex == "F"]) / 1000, 1)

mean_harvest_m <- round(mean(d_ind$harvest[d_ind$sex == "M"]), 1)
mean_harvest_f <- round(mean(d_ind$harvest[d_ind$sex == "F"]), 1)

pooled <- sum(d_ind_pooled$pooled == "Pooled")

zeros_m <- 100*round(sum(d_ind$harvest[d_ind$sex == "M"] == 0) / sum(d_ind$sex == "M"), 3)
zeros_f <- 100*round(sum(d_ind$harvest[d_ind$sex == "F"] == 0) / sum(d_ind$sex == "F"), 3)
```

There were `r nrow(d)` hunting trips, `r pct_solo`% of which involved solo hunters, and the remaining mostly groups of 2 or 3. There was possibly a sex bias in data collection: most of the studies reporting female hunting (6/7) had a female coauthor, although many studies with female coauthors did not report female hunting (16/22):

```{r}
tbl_coauthor
summary(tbl_coauthor)
```

Keeping in mind the potential bias in data collection, there were pronounced sex differences in these data. The breakdown of index hunters by sex was `r males` males (`r pct_males`%) and `r females` females (`r pct_females`%), for a total of `r nrow(hunters)` index hunters (this is less than the total hunters, which also included assistants).

This was the distribution of index hunter ages by sex:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(hunters, aes(Age)) +
  geom_histogram() +
  facet_wrap(~Sex, ncol = 1, scales = 'free_y') +
  theme_bw(15)
```

To assess sex differences in hunting returns (harvests), I removed `r mixed` hunting trips from the data (`r pct_mixed`%) that included hunting parties with both males and females. Males returned a total of `r harvest_males` metric tons of meat, and females `r harvest_females` metric tons. Removing hunting trips with pooled harvests (N=`r pooled`), the mean hunting return for men was `r mean_harvest_m` kg, and for women `r mean_harvest_f` kg. However, men returned empty handed in `r zeros_m`% of trips, and women in only `r zeros_f`% of trips.

Finally, here is the distribution of returns by sex and society:

![Returns of subsistence hunting trips (N=23,652) by 1,518 members of 40 forager and agricultural societies. Row labels refer to the group and country from which the data were collected (some studies did not include group names). Red labels indicate studies with one or more female authors. Points indicate returns of hunting trips (in kilograms). Point size indicates the number of returns of that mass. X-axis is on a log scale (1 kg is added to the mass of each return so that a return of 0, i.e., returning empty-handed, is well-defined on a log scale). Rows are sorted by the mean return for each society. Individual: returns are attributable to individual hunters. Pooled: returns cannot be attributed to individual hunters. Data from Koster et al. (2020). It is important to note that these data are not necessarily representative of hunting patterns in foraging groups, and that there are certainly biases](plot_koster.svg)
